P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2024-06-29 manuscript.tex
minor
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

Many performance-oriented researches exist for MD systems, most of them using MD packages as a black-box and adjusting different simulation parameters to analyse their implications in the overall performance.
In some cases, however, in-depth performance studies were made to gain a broader understanding on efficient ways to achieve SIMD vectorization with SSE/AVX instructions \cite{mdsimd}. Most in-depth studies are performed using mini-apps or proxy-apps, with miniMD being worth to mention as it provides a simple copper face-cubic-centered lattice simulation using the kernels extracted from LAMMPS.

@@

Based on our searches, there are not many in-depth performance studies focusing on SIMD and in-core execution for kernels used in DEM simulations, and most existing studies focus on evaluating their distributed-memory parallelism in weak-scaling settings.
With P4IRS, we intend to bridge the gap between performance strategies across different simulation domains, and use the most efficient algorithm whenever it is possible, without requiring significant efforts such as re-implementing kernels and re-designing applications with tailored data structures for improving performance.

@@

Particle simulations consist of solving N-body problems and have a wide range of applications as many fields can be represented as a system of interacting particles.
In this sense, particles can be either atoms, rigid bodies or even planets, stars and galaxies.
Interaction contributions can come either from inter-atomic van-der-Walls forces based on the Lennard-Jones potential, or even from gravitational forces generated by super-massive particles.
In order to simulate such systems, however, we need well-defined and closed system of particles with known boundary conditions.
In the scope of this paper, particle contributions are computed for particles that lie within a range or ``cutoff radius''. 
Strategies to compute approximations from particles beyond this range (also known as long-range contributions) also exist, but are not disclosed.
To provide distinct examples of simulations and display a certain level of flexibility from P4IRS, we focus on MD and DEM simulations in this work.

@@

With the torque vector $\hat{\tau}{_i}$, angular velocity vector $\hat{\omega}{_i}$, rotation $\varphi$ (typically stored as a quaternion, rotation matrix or rotation vector) and moment of inertia about the center of mass $I = I(\varphi_i)$.
A collision detection step is required to check when two particles overlap, and parameters about the collision such as the contact point and penetration depth are calculated and used in the force calculation.
Note that the reason a penetration occurs during a collision is that the collision is not detected right when it starts because the simulation is discretized by the time-step size.
For simplification purposes, we only calculate collision detections analytically in this work since there are no particles with complex shapes involved.

@@

Despite avoiding computations from long-distance particles, another common optimization is to take advantage of Newton's Third Law and computing forces in only one direction.
This means that only half of particle-pairs are traversed during force computation, and forces are decreased for the neighbor particles during their computation.
Ideally, this should lead to a factor of two speedup, but this strategy has a few drawbacks: (a) parallelism is harmed since decreasing force of neighbor particles requires atomic operations to avoid race conditions and (b) when the code is vectorized with AVX/SSE intrinsics, gather and scatter instructions are needed to build up the SIMD registers with forces of the neighbor particles, which results in scattered memory accesses and extra instructions.
At the end, this strategy is likely to be beneficial for more arithmetic-intensive kernels, and the trade-off between parallelism and less arithmetic operations must be assessed.

@@

P4IRS\footnote{\url{https://i10git.cs.fau.de/software/pairs}} is an open-source, stand-alone compiler and domain-specific language for particle simulations which aims at generating optimized code for different target hardwares.
It is released as a Python package and allows users to define kernels, integrators and other particle routines in a high-level and straightforward fashion.

@@

\autoref{lst:pairslj} shows a simple Lennard-Jones kernel implemented in P4IRS.
Functions can be provided through a method (line 1), and the parameters in the method refer to different particles in the system.
In this way, many-body potentials can also be described using methods with more than two parameters.
Some keywords are also provided by P4IRS for writing the kernels, in line 2 the \emph{squared\_distance} method is used for obtaining the squared distance for the atoms \emph{i} and \emph{j}, and in line 4 the \emph{delta} method is used for obtaining the delta vector.
Also, in lines 3 and 4 it is possible to see accesses to feature properties called \emph{sigma6} and \emph{epsilon}, which are defined during the simulation setup as properties which depend on the atom types feature.
In line 4, the \emph{apply} method is used for applying the computed term into the \emph{force} property of the particles, the method allows the identification of particle properties reductions and is also used when generating the kernel variant using the half neighbor-lists.

@@

%\begin{lstlisting}[language=Python,
%		   label={lst:pairslj},
%		   caption={Lennard-Jones force description in P4IRS.}]
%\end{lstlisting}

@@

Apart from the kernel implementations and functions to prepare the kernel launches and move data between host and accelerator processors, particle simulations also may need to call external libraries and deal with filesystem I/O.
A few case examples where this is needed are, for instance, to write and read particle data into/from a visualization format like VTK, and reading the initial state of the particle system to setup the simulation.
Since most of these procedures do not require extensive optimizations as they are not compute-intensive and usually their bottleneck is concentrated in an external I/O device (such as disks), building up the intermediate representation for them and extending the code generator to support such external calls would demand significant effort with no substantial benefit.

@@

\autoref{lst:pairsruntime} shows an example for the routine to write data into VTK files, note that the first parameter is a \emph{PairsRuntime} object, and can be used to check whether properties exist, check if they are synced with the host device and (in case not) copy the data from the host.
The meta-data object is feed at the start of the simulation with the code generated by P4IRS, and it is therefore dependent on each simulation.
It is important to enhance that accessing simulation data through the \emph{PairsRuntime} instance is not as efficient as accessing them directly in the generated code since the way of accessing the data must be evaluated during runtime.
Therefore, the usage of such instance must be avoided for performance-intensive routines.

@@

For \ac{MD} case, we use the Verlet Lists algorithm as it is the standard one for such simulations.
The simulations runs through 200 time-steps, and the neighbor-lists are rebuilt at every 20 time-steps of the simulation (and at the start).
We use the simple Lennard-Jones potential (as shown in \autoref{lst:pairslj}) to calculate the force contributions among atoms.

>>>>

Many performance-oriented researches exist for MD systems, most of them using MD packages as a black-box and adjusting different simulation parameters to analyse their implications in the overall performance.
In some cases, however, in-depth performance studies were made to gain a broader understanding on efficient ways to achieve SIMD vectorization with SSE/AVX instructions \cite{mdsimd}.
Most in-depth studies are performed using mini-apps or proxy-apps, one example is miniMD, a proxy-app that provides a simple copper face-cubic-centered lattice simulation where atom forces are computed with kernels extracted from LAMMPS for either the \ac{LJ} or \ac{EAM} potentials.

@@

Based on our searches, there are not many in-depth performance studies focusing on SIMD and in-core execution for kernels used in DEM simulations, and most existing studies focus on evaluating their distributed-memory parallelism in weak-scaling settings.
With P4IRS, we intend to bridge the gap between performance strategies across different simulation domains, and use the most efficient algorithm whenever it is possible without requiring significant efforts such as re-implementing kernels and re-designing applications with tailored data structures for improving performance.

@@

Particle simulations consist of solving N-body problems and have a wide range of applications as many fields can be represented as a system of interacting particles.
In this sense, particles can be either atoms, rigid bodies or even planets, stars and galaxies.
Interaction contributions can come either from inter-atomic van-der-Walls forces based on the \ac{LJ} potential, or even from gravitational forces generated by super-massive particles.
In order to simulate such systems, however, we need well-defined and closed system of particles with known boundary conditions.
In the scope of this paper, particle contributions are computed for particles that lie within a range or ``cutoff radius''. 
Strategies to compute approximations from particles beyond this range (also known as long-range contributions) also exist, but are not disclosed.
To provide distinct examples of simulations and display a certain level of flexibility from P4IRS, we focus on MD and DEM simulations in this work.

@@

With the torque vector $\hat{\tau}{_i}$, angular velocity vector $\hat{\omega}{_i}$, rotation $\varphi$ (typically stored as a quaternion, rotation matrix or rotation vector) and moment of inertia about the center of mass $I = I(\varphi_i)$.
A collision detection step is required to check when two particles overlap, and parameters about the collision such as the contact point and penetration depth are computed and used in the force calculation.
Note that the reason a penetration occurs during a collision is that the collision is not detected right when it starts because the simulation is discretized by the time-step size.
For simplification purposes, we only calculate collision detections analytically in this work since there are no particles with complex shapes involved.

@@

Despite avoiding computations from long-distance particles, another common optimization is to take advantage of Newton's Third Law and compute forces in only one direction.
This means that only half of particle-pairs are traversed when computing the forces, and partial forces are decreased for the neighbor particles during their computation.
Ideally, this should lead to a factor of two speedup, but this strategy has a few drawbacks: (a) parallelism is harmed since decreasing force of neighbor particles requires atomic operations to avoid race conditions and (b) when the code is vectorized with SIMD intrinsics (such as AVX2 and AVX512), gather and scatter instructions are needed to build up the SIMD registers with forces of the neighbor particles, which results in scattered memory accesses and extra instructions.
At the end, this strategy is likely to be beneficial for more arithmetic-intensive kernels, and the trade-off between parallelism and less arithmetic operations must be assessed to know when it is advantageous.

@@

P4IRS\footnote{\url{https://i10git.cs.fau.de/software/pairs}} is an open-source, stand-alone compiler and domain-specific language for particle simulations which aims at generating optimized code for different target hardwares.
It is released as a Python package and allows users to define kernels, integrators and other particle routines in a high-level and straightforward fashion without the need to implement any backend code.

@@

\autoref{lst:pairslj} shows a simple description in P4IRS to compute forces with the \ac{LJ} potential.
Functions can be provided through a method (line 1), and the parameters in the method refer to different particles in the system.
In this way, many-body potentials can also be described using methods with more than two parameters.
Some keywords are also provided by P4IRS for writing the kernels, in line 2 the \emph{squared\_distance} method is used for obtaining the squared distance for the atoms \emph{i} and \emph{j}, and in line 4 the \emph{delta} method is used for obtaining the delta vector.
Also, in lines 3 and 4 it is possible to see accesses to feature properties called \emph{sigma6} and \emph{epsilon}, which are defined during the simulation setup as properties which depend on the atom types feature.
In line 4, the \emph{apply} method is used for applying the computed term into the \emph{force} property of the particles, the method allows the identification of particle properties reductions and is also used when generating the kernel variant using the half neighbor-lists.

@@

@@

Apart from the kernel implementations and functions to prepare the kernel launches and move data between host and accelerator processors, particle simulations also may need to call external libraries and deal with filesystem I/O.
A few case examples where this is needed are, for instance, to write particle data to a file in a visualization format like VTK, and reading the initial state of the particle system to setup the simulation.
Since most of these procedures do not require extensive optimizations as they are not compute-intensive and usually their bottleneck is concentrated in an external I/O device (such as disks), building up the intermediate representation for them and extending the code generator to support such external calls would demand significant effort with no substantial benefit.

@@

\autoref{lst:pairsruntime} shows an example for the routine to write data into VTK files, note that the first parameter is a \emph{PairsRuntime} object, and can be used to check whether properties exist, check if they are synced with the host device and (in case not) copy the data from the host.
The meta-data object is feed at the start of the simulation with the code generated by P4IRS, and it is therefore dependent on each simulation.
It is important to enhance that accessing simulation data through the \emph{PairsRuntime} instance is not as efficient as accessing them directly in the generated code since the way of accessing the data structures (such as the property layouts) must be evaluated during runtime.
Therefore, the usage of such instance must be avoided for performance-intensive routines.

@@

For \ac{MD} case, we use the Verlet Lists algorithm as it is the standard one for such simulations.
The simulations runs through 200 time-steps, and the neighbor-lists are rebuilt at every 20 time-steps of the simulation (and at the start).
We use the simple \ac{LJ} potential (as shown in \autoref{lst:pairslj}) to calculate the force contributions among atoms.
